---
title: "Plan372 HW 2"
author: "Annie Georgeson"
format: html
editor: visual
---

Load the necessary libraries

```{r}
library(tidyverse)
library(lubridate)
```

Load the data from CSV.

```{r}
data = read.csv(here::here("Fire_Incidents.csv"))
```

Extract the first few rows and all columns to see what you're looking at.

```{r}
head(data)
```

There are 20 columns

X, Y – longitude and latitude of incident

OBJECTID, incident_number – numeric identifiers for each incident

incident_type – numeric code for the type of the incident (e.g. house fire, smell of gas, etc)

incident_type_description – plain-English description of the incident type code

arrive_date_time – when firefighters arrived at the scene of the incident

dispatch_date_time – when the call came in and firefighters were dispatched to the incident

cleared_date_time – when firefighters cleared the incident and left the scene

exposure – if the fire spread from another fire (e.g. a house fire started by a car fire)

platoon – which shift responded to the incident

station – which fire station responded to the incident

address, address2 – address and address line 2 for the incident

apt_room – apartment or suite number, if applicable

GlobalID – numeric identifier

CreationDate, EditDate – when the incident was entered into the system, and when the record was last edited

Creator, Editor – who entered the incident, and who edited it most recently

# 1) Average time it takes Wake County to respond to incidents

```{r}
# Convert the date-time columns 
data$dispatch_date_time <- ymd_hms(data$dispatch_date_time)
data$arrive_date_time <- ymd_hms(data$arrive_date_time)

# Calculate response time in seconds
data$response_time <- as.numeric(difftime(data$arrive_date_time, data$dispatch_date_time, units = "secs"))

# Calculate the average response time in minutes
average_response_time_minutes <- mean(data$response_time, na.rm = TRUE) / 60

# Display the average response time in minutes
average_response_time_minutes

```

#### The average response time is approximately 5.3 minutes.

# 2) Varying response times by different stations

```{r}
# Group by station and calculate average response time
library(dplyr)

average_response_by_station <- data |>
  group_by(station) |>
  summarise(mean_response_time = mean(response_time, na.rm = TRUE) / 60)

# Display stations with the highest and lowest response times
arrange(average_response_by_station, mean_response_time)

```

#### Station 13 has the lowest average response time & Station 29 has the highest average response time.

# 3) Trend of Wake County Fire's response time over time

```{r}
# Extract the year from the dispatch date-time
data$year <- year(data$dispatch_date_time)

# Group by year and calculate average response time per year
response_time_by_year <- data |>
  group_by(year) |>
  summarise(mean_response_time = mean(response_time, na.rm = TRUE) / 60)

# Display the trend on a plot
ggplot(response_time_by_year, aes(x=year, y=mean_response_time)) +
  geom_line()

```

#### The response times for Wake County Fire have stayed relatively consistent with spikes in response time being seen towards the end of 2018 and beginning of 2019.

#### Possible reasons for the spikes at these points in time is that the distance that Wake County Fire had to travel might have increased or there might have been a greater amount of incidents that led to a delay in response time.

# Time of day that calls are most likely to occur

```{r}
# Extract hour from the dispatch date-time
data$hour = hour(data$dispatch_date_time)

# Count incidents by hour
incidents_by_hour = data |>
  group_by(hour) |>
  summarize(count=n())

# Display the distributions of incidents by hour
incidents_by_hour

```

#### Calls are most likely to occur at the times of 10pm, 9pm, 11pm, and 6pm.

# Amount of calls to Wake County Fire in this dataset, and how many of them are actual fires

```{r}
# display the total number of calls in the dataset 
total_calls = nrow(data)
print(total_calls)
```

```{r}
# filter for fire incidents
fire_incidents = data |>
  filter(grepl("Fire", incident_type_description, ignore.case = TRUE))

# display the number of fire incidents
num_fire_incidents = nrow(fire_incidents)
print(num_fire_incidents)
```

#### There was a total of 229,047 calls in the data set and 51,976 of those were calls for fire incidents.

# Average response time to actual fires

```{r}
# Calculate the average response time for fire incidents 
average_fire_response_time = mean(fire_incidents$response_time, na.rm = TRUE) / 60

average_fire_response_time
```

#### The average response time for fire incidents was approximately 4.9 minutes, which is faster than the overall response time for all incidents.

# 
